---
title: "Explore clusters"
author: "Rose Hartman"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
params:
  n_cluster: 19
  cluster: "group19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      out.width = "50%")

library(ggplot2)

# load custom functions for plotting
source(here::here("src", "scripts", "functions_plotting.R"))

# if the figures directory doesn't exist, create it
dir.create(here::here("reports"), showWarnings = FALSE)
dir.create(here::here("reports", "figures"), showWarnings = FALSE)
dir.create(here::here("reports", "tables"), showWarnings = FALSE)
```

`r if(!knitr::opts_chunk$get()$echo) ">Note that code chunks are not printed in this report in order to keep the output tidy. To see all of the code to generate these results, open the .Rmd file."`


```{r reformat_data}
needs_assessment_long <- readRDS(here::here("data", "deidentified", "needs_assessment_with_clusters_w2.rds")) |> 
  # reformat to long
  tidyr::pivot_longer(cols = dplyr::matches("_relevance|_expertise|_learn")) |> 
  # separate topic (e.g. data_viz) from question type (relevance, expertise, or learn)
  tidyr::extract(name, into = c("topic", "question"), regex = "(.*)_([[:lower:]]+)", remove = TRUE) |> 
  dplyr::mutate(question = factor(question, levels = c("relevance", "expertise", "learn")),
                topic = as.factor(topic))
```

# Solution with `r params$n_cluster` clusters

## distributions by cluster 

```{r cluster_distributions, eval = FALSE}
plot_by_theme(needs_assessment_long, cluster=(!!sym(params$cluster)))
```

## Which topics are most important to each cluster?

for each cluster, rank topics by desire to learn, expertise, or relevance

```{r rank_topics}
rank_topics(needs_assessment_long, cluster = (!!sym(params$cluster)), n_cluster = params$n_cluster)
```
